Accessible, realistic genome simulation with selection using \stdpopsim

### 2025-03-17 analysis2_manuscript.tex
discussion of AUC scores for sweeps

     %sweeep power figure
    \begin{figure}[t!]
        \centering
        \includegraphics[width=0.8 \textwidth]{figures/sweeps/chr1_power.pdf}
        \caption{
        Power to detect selective sweeps at 100 locations along human
        chromosome 1. Genomic segments were simulated with sweeps under a three population out-of-Africa model
        \citep{gutenkunst2009inferring} and with background selection from deleterious mutations in exons.
        %IG: seems like methods-level detail
        %Single beneficial mutations were introduced at each location with a selection
        % coefficient of $s = 0.03$, and only sampled if they reached a terminal
        % frequency of 0.95 or greater.
        Three methods for detecting sweeps were applied to simulated data:
        1) \sweepfinder \citep{degiorgio2016sweepfinder2} (top row),
        2) \diploshic \citep{kern2018diplos} (middle row),
        and 3) reduced diversity ($\pi$) (bottom row).
        Power (true positive rate) is shown for these methods for the CEU and YRI
        samples (left and right respectively).
        The thresholds of the test statistics were set to control for
        $5\%$ false positive rate under a neutral null model (blue)
        and a null model with background selection from deleterious mutations in exons (red).
        Also shown are heatmaps of exon density and local recombination rate along the
        chromosome (bottom).
        \igcomment{In fig, replace CLR with \sweepfinder; also in figs 6 S5 S6}
        }
        \label{fig:chr1_power}
    \end{figure}

@@

    We compared the overall performance of the three sweep-detection methods using
    receiver operating characteristic (ROC) curves (Figure \ref{fig:roc-curves}).
    Different methods perform better for each of the sampled populations,
    with \sweepfinder performing better than the other two on the CEU sample,
    and \diploshic performing better on the YRI sample.
    This is may result from the fact that we trained \diploshic on simulations under a constant size population,
    which more closely matches the demographic history of YRI, than that of CEU, which 
    is characterized by a strong recent bottleneck followed by population expansion.
    %\drscomment{note: I am couching the language above a bit because we have previously shown that shic doesn't suffer in the sweep vs. nonsweep question when doing other train-on-constant-test-on-nonequilib benchmarks. Could be something else going on here (or some interaction between demog misspec and rec rate variation), but in any case I am not certain of the cause so I am toning down our certainty slightly.}
    Interestingly, the simple method that uses reduced diversity ($\pi$) as an indicator
    for a selective sweep performs reasonably well overall (especially on the CEU sample).
    However, its accuracy is much more variable along the chromosome compared to the other two method (see Figure \ref{fig:chr1_power}),
    and for quite a few locations along the chromosome it has power below $50\%$
    when applied to both sample sets.
    \sweepfinder appears to have the least variation in power along the chromosome or across populations.

    \begin{figure}[b!]
        \centering
        \includegraphics[width=0.8 \textwidth]{figures/sweeps/relationship_power_cM.pdf}
        \caption{
        Power to detect selective sweeps 
        %IG: shortened the title line
        %at 100 locations along human
        %chromosome 1 under a three population out-of-Africa model
        %\citep{gutenkunst2009inferring}  when using null models with and
        %without purifying selection
        as a function of local recombination rate.
        This figure shows the same power estimates shown in Figure \ref{fig:chr1_power},
        but with the 100 genomic segments plotted against their
        average recombination rates instead of position along chromosome 1.
    	Genomic segments were simulated with sweeps under a three population out-of-Africa model
        \citep{gutenkunst2009inferring} and with background selection from deleterious mutations in exons.
        Three methods for detecting sweeps were applied to simulated data:
        1) \sweepfinder \citep{degiorgio2016sweepfinder2} (top row),
        2) \diploshic \citep{kern2018diplos} (middle row),
        and 3) reduced diversity ($\pi$) (bottom row).
        Power (true positive rate) is shown for these methods for the CEU and YRI
        samples (left and right respectively).
        The thresholds of the test statistics were set to obtain a
        $5\%$ false positive rate under a neutral null model (blue)
        and a null model with background selection from deleterious mutations in exons (red).
        Fitted lines represent loess smoothed regressions.
        }
        \label{fig:power-recomb}
    \end{figure}

@@

% roc curves for sweeps
\begin{figure}
    \centering
    \includegraphics[width=0.8 \textwidth]{figures/sweeps/roc_neutral_null.pdf}
    \caption{
    \label{fig:roc-curves}
    Receiver operating characteristic (ROC) curves for the three methods for detecting sweeps.
    True positive rates (TPR) were computed across all 100,000 datasets simulated with selective sweeps,
    and false positive rates (FPR) were established by applying the methods to genomes simulated
    under neutral evolution with the same demographic history and genetic map
    used in the simulations that included sweeps (see \textbf{Methods}).
    \igcomment{Add AUCs?}
    }
    %DRS-fig: can we include AUCs here?
\end{figure}

%sweep power vs exon density
\begin{figure}
    \centering
    \includegraphics[width=0.8 \textwidth]{figures/sweeps/relationship_power_exon.pdf}
    \caption{
    %IG: copied new caption from fig 6 and adjusted.
        Power to detect selective sweeps as a function of local exon density.
        This figure shows the same power estimates shown in Figure \ref{fig:chr1_power},
        but with the 100 genomic segments sorted according to their
        exon density (percentage of exonic bases per window)instead of position along chromosome 1.
	    Genomic segments were simulated with sweeps under a three population out-of-Africa model
        \citep{gutenkunst2009inferring} and with background selection from deleterious mutations in exons.
        Three methods for detecting sweeps were applied to simulated data:
        1) \sweepfinder \citep{degiorgio2016sweepfinder2} (top row),
        2) \diploshic \citep{kern2018diplos} (middle row),
        and 3) reduced diversity ($\pi$) (bottom row).
        Power (true positive rate) is shown for these methods for the CEU and YRI
        samples (left and right respectively).
        The thresholds of the test statistics were set to obtain a
        $5\%$ false positive rate under a neutral null model (blue)
        and a null model with background selection from deleterious mutations in exons (red).
        Fitted lines represent loess smoothed regressions.
    }
    \label{fig:power-exon}
\end{figure}

>>>>

     %sweeep power figure
    \begin{figure}[t!]
        \centering
        \includegraphics[width=0.8 \textwidth]{figures/sweeps/chr1_power.pdf}
        \caption{
        Power to detect selective sweeps at 100 locations along human
        chromosome 1. Genomic segments were simulated with sweeps under a three population out-of-Africa model
        \citep{gutenkunst2009inferring} and with background selection from deleterious mutations in exons.
        %IG: seems like methods-level detail
        %Single beneficial mutations were introduced at each location with a selection
        % coefficient of $s = 0.03$, and only sampled if they reached a terminal
        % frequency of 0.95 or greater.
        Three methods for detecting sweeps were applied to simulated data:
        1) \sweepfinder \citep{degiorgio2016sweepfinder2} (top row -- labelled CLR),
        2) \diploshic \citep{kern2018diplos} (middle row),
        and 3) reduced diversity ($\pi$) (bottom row).
        Power (true positive rate) is shown for these methods for the CEU and YRI
        samples (left and right respectively).
        The thresholds of the test statistics were set to control for
        $5\%$ false positive rate under a neutral null model (blue)
        and a null model with background selection from deleterious mutations in exons (red).
        Also shown are heatmaps of exon density and local recombination rate along the
        chromosome (bottom).
        }
        \label{fig:chr1_power}
    \end{figure}

@@

    We compared the overall performance of the three sweep-detection methods using
    receiver operating characteristic (ROC) curves (Figure \ref{fig:roc-curves}).
    Performance varies for each of the sampled populations.
    In the CEU sample, \sweepfinder and \diploshic perform similarly with a high area under the curve (AUC) score,
    (AUC = 0.906 and 0.903 respectively), but both underperform the simple method
    that uses reduced diversity ($\pi$) as an indicator (AUC = 0.932).
    In the YRI sample, \diploshic outperforms the other two methods,
    with an AUC of 0.979, followed by using outliers of $\pi$ with an AUC of 0.957,
    and then \sweepfinder with an AUC of 0.939.
    We note that we trained \diploshic on simulations under a constant size population,
    which more closely matches the demographic history of YRI, than that of CEU, which 
    is characterized by a strong recent bottleneck followed by population expansion.
    %\drscomment{note: I am couching the language above a bit because we have previously shown that shic doesn't suffer in the sweep vs. nonsweep question when doing other train-on-constant-test-on-nonequilib benchmarks. Could be something else going on here (or some interaction between demog misspec and rec rate variation), but in any case I am not certain of the cause so I am toning down our certainty slightly.}
    Interestingly, the simple method that uses reduced diversity ($\pi$) as an indicator
    for a selective sweep performs very well overall (especially on the CEU sample).
    However, its accuracy is much more variable along the chromosome compared to the other two method (see Figure \ref{fig:chr1_power}),
    and for quite a few locations along the chromosome it has power below $50\%$
    when applied to both sample sets.
    \sweepfinder appears to have the least variation in power along the chromosome or across populations.

    \begin{figure}[b!]
        \centering
        \includegraphics[width=0.8 \textwidth]{figures/sweeps/relationship_power_cM.pdf}
        \caption{
        Power to detect selective sweeps 
        %IG: shortened the title line
        %at 100 locations along human
        %chromosome 1 under a three population out-of-Africa model
        %\citep{gutenkunst2009inferring}  when using null models with and
        %without purifying selection
        as a function of local recombination rate.
        This figure shows the same power estimates shown in Figure \ref{fig:chr1_power},
        but with the 100 genomic segments plotted against their
        average recombination rates instead of position along chromosome 1.
    	Genomic segments were simulated with sweeps under a three population out-of-Africa model
        \citep{gutenkunst2009inferring} and with background selection from deleterious mutations in exons.
        Three methods for detecting sweeps were applied to simulated data:
        1) \sweepfinder \citep{degiorgio2016sweepfinder2} (top row-- labelled CLR),
        2) \diploshic \citep{kern2018diplos} (middle row),
        and 3) reduced diversity ($\pi$) (bottom row).
        Power (true positive rate) is shown for these methods for the CEU and YRI
        samples (left and right respectively).
        The thresholds of the test statistics were set to obtain a
        $5\%$ false positive rate under a neutral null model (blue)
        and a null model with background selection from deleterious mutations in exons (red).
        Fitted lines represent loess smoothed regressions.
        }
        \label{fig:power-recomb}
    \end{figure}

@@

% roc curves for sweeps
\begin{figure}
    \centering
    \includegraphics[width=0.8 \textwidth]{figures/sweeps/roc_neutral_null.pdf}
    \caption{
    \label{fig:roc-curves}
    Receiver operating characteristic (ROC) curves for the three methods for detecting sweeps.
    True positive rates (TPR) were computed across all 100,000 datasets simulated with selective sweeps,
    and false positive rates (FPR) were established by applying the methods to genomes simulated
    under neutral evolution with the same demographic history and genetic map
    used in the simulations that included sweeps (see \textbf{Methods}).
    AUC scores are shown for each method. \sweepfinder is labelled CLR in the figure.
    }
    %DRS-fig: can we include AUCs here?
\end{figure}

%sweep power vs exon density
\begin{figure}
    \centering
    \includegraphics[width=0.8 \textwidth]{figures/sweeps/relationship_power_exon.pdf}
    \caption{
    %IG: copied new caption from fig 6 and adjusted.
        Power to detect selective sweeps as a function of local exon density.
        This figure shows the same power estimates shown in Figure \ref{fig:chr1_power},
        but with the 100 genomic segments sorted according to their
        exon density (percentage of exonic bases per window)instead of position along chromosome 1.
	    Genomic segments were simulated with sweeps under a three population out-of-Africa model
        \citep{gutenkunst2009inferring} and with background selection from deleterious mutations in exons.
        Three methods for detecting sweeps were applied to simulated data:
        1) \sweepfinder \citep{degiorgio2016sweepfinder2} (top row -- labelled CLR),
        2) \diploshic \citep{kern2018diplos} (middle row),
        and 3) reduced diversity ($\pi$) (bottom row).
        Power (true positive rate) is shown for these methods for the CEU and YRI
        samples (left and right respectively).
        The thresholds of the test statistics were set to obtain a
        $5\%$ false positive rate under a neutral null model (blue)
        and a null model with background selection from deleterious mutations in exons (red).
        Fitted lines represent loess smoothed regressions.
    }
    \label{fig:power-exon}
\end{figure}
